Xolotl: An Intuitive and Approachable Neuron and Network Simulator for Research and Teaching

### 2018-08-17 paper/paper.tex
another minor typo

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Xolotl allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parts of the model. The simulator's architecture allows for the interactive manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Xolotl is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Its ease of use and rich visualization capabilities make it an attractive option in teaching environments. Finally, xolotl is written in a modular fashion, includes detailed tutorials and worked examples, and is freely available at https://github.com/sg-s/xolotl, enabling seamless integration into the workflow of other researchers.

@@

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we will show how xolotl can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each "conductance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum. In summary, xolotl can be used to construct neuron models with intracellular mechanisms such as Calcium influx and buffering, and homeostatic regulation.

@@

Network models in xolotl consist of "compartment" objects that can be connected by "synapse" objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, to connect two single-compartment neurons called "LP" and "PY" using a chemical synapse of type "Cholinergic" with strength 30 nS, we can use

>>>>

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Xolotl allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parts of the model. The simulator's architecture allows for the interactive manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Xolotl is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Its ease of use and rich visualization capabilities make it an attractive option in teaching environments. Finally, xolotl is written in a modular fashion, includes detailed tutorials and worked examples, and is freely available at https://github.com/sg-s/xolotl, enabling seamless integration into the workflows of other researchers.

@@

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we will show how xolotl can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each conductance object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum. In summary, xolotl can be used to construct neuron models with intracellular mechanisms such as Calcium influx and buffering, and homeostatic regulation.

@@

Network models in xolotl consist of compartment objects that can be connected by synapse objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, to connect two single-compartment neurons called "LP" and "PY" using a chemical synapse of type "Cholinergic" with strength 30 nS, we can use
